---
title: "Precision communication: Physicians’ linguistic adaptation to patients’ health literacy"
author: "Nick Duran"
date: 10/14/21
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
```

## What this Notebook Contains

> Assuming access to the available source data files, the Notebook contains the complete statistical code for replicating:
> 
> * "Table 1. Characteristics of patients in overall sample"

#### Step 1. Import "Duran_Schillinger_KNPC_restricted_demographics.csv."

See README in repository for instructions on how to access the .csv file. 

**Please note:** *Kaiser Permanante (owners of data) will not allow any individual-level data (even a small subset) to be shown in a public repository. A version of this HTML exists with the restricted .csv files that includes a subset of individual-level data to aid in tutorial interpretation.*

```{r, message=FALSE}
library(tidyverse)
library(pander)

df.demo = read.csv("Duran_Schillinger_KNPC_restricted_demographics.csv")
pander(names(df.demo))
```

- **PAT_ID**: Randomly generated numeric code associated with each unique patient 
- **Age**: Age of patient
- **sex**: Sex of patient; 0 = female, 1 = male
- **race**: Race of patient; 1 = White, 2 = Black, 3 = Hispanic, 4 = Asian, 5 = Other 
- **educ_p**: Education level of patient; 1 = No degree, 2 = GED/high school, 3 = Some college or more    
- **CHARLSON_INDEX**: Co-morbidity index, interval
- **predictions_LP**: Patient's assigned score of "high" or "low" health literacy (otherwise known as "LP Score") based on previously developed linguistic model; see: S. A. Crossley, R. Balyan, J. Liu, A. J. Karter, D. McNamara, D. Schillinger, Developing and testing automatic models of patient communicative health literacy using linguistic features: Findings from the ECLIPPSE study, Health Commun. (2020)

#### Step 2: Reassign variable levels for readability

```{r}
df.demo.prep = df.demo %>%
  mutate(sex = case_when(
    sex=="0" ~ "Women", 
    sex=="1" ~ "Men",
    TRUE ~ NA_character_)) %>%
  mutate(race = case_when(
    race=="1" ~ "White", 
    race=="2" ~ "Black",
    race=="3" ~ "Hispanic", 
    race=="4" ~ "Asian",
    race=="5" ~ "Other", 
    TRUE ~ NA_character_)) %>%
  mutate(educ_p = case_when(
    educ_p=="1" ~ "No degree", 
    educ_p=="2" ~ "GED/high school",
    educ_p=="3" ~ "Some college or more",    
    TRUE ~ NA_character_)) 

```

#### Step 3: For each variable, generate summary statistics based on whether high or low health literacy; Start with education level as example

```{r, warning=FALSE}

## Get stratified counts and percentage of each education category
demo_ed = df.demo.prep %>% group_by(educ_p) %>%
  summarize(count=n(), perce = (n()/nrow(df.demo.prep))*100) 
demo_edlow = filter(df.demo.prep, predictions_LP=="Low") %>% group_by(educ_p) %>%
  summarize(count=n(), perce = (n()/nrow(filter(., predictions_LP=="Low")))*100)  
demo_edhigh = filter(df.demo.prep, predictions_LP=="High") %>% group_by(educ_p) %>%
  summarize(count=n(), perce = (n()/nrow(filter(., predictions_LP=="High")))*100)  

## Assemble into readable matrix
demoEduc = matrix(c(
  "No degree ", demo_ed$count[2], round(demo_ed$perce[2],1) , demo_edlow$count[2], round(demo_edlow$perce[2],1), demo_edhigh$count[2], round(demo_edhigh$perce[2],1),  
  "GED/high school", demo_ed$count[1], round(demo_ed$perce[1],1) , demo_edlow$count[1], round(demo_edlow$perce[1],1), demo_edhigh$count[1], round(demo_edhigh$perce[1],1),  
  "Some college or more", demo_ed$count[3], round(demo_ed$perce[3],1) , demo_edlow$count[3], round(demo_edlow$perce[3],1), demo_edhigh$count[3], round(demo_edhigh$perce[3],1) ),  
  ncol=7,byrow=TRUE)
colnames(demoEduc) = c("Levels", "Total", "Total-%", "LowHL", "LowHL-%", "HighHL", "HighHL-%")

pander(head(demoEduc))
```

#### Step 4: Follow same procedure as above for remaining variables

```{r}

## Get stratified counts and percentage of each sex category
demo_sex = df.demo.prep %>% group_by(sex) %>%
  summarize(count=n(), perce = (n()/nrow(df.demo.prep))*100)  
demo_sexlow = filter(df.demo.prep, predictions_LP=="Low") %>% group_by(sex) %>%
  summarize(count=n(), perce = (n()/nrow(filter(., predictions_LP=="Low")))*100)  
demo_sexhigh = filter(df.demo.prep, predictions_LP=="High") %>% group_by(sex) %>%
  summarize(count=n(), perce = (n()/nrow(filter(., predictions_LP=="High")))*100)  

## Assemble into readable matrix
demoSex = matrix(c(
  "Women", demo_sex$count[2], round(demo_sex$perce[2],1) , demo_sexlow$count[2], round(demo_sexlow$perce[2],1), demo_sexhigh$count[2], round(demo_sexhigh$perce[2],1) ),
  ncol=7,byrow=TRUE)
colnames(demoSex) = c("Levels", "Total", "Total-%", "LowHL", "LowHL-%", "HighHL", "HighHL-%")

pander(demoSex)
```

```{r}

## Get stratified counts and percentage of each race category
demo_race = df.demo.prep %>% group_by(race) %>%
  summarize(count=n(), perce = (n()/nrow(df.demo.prep))*100) 
demo_racelow = filter(df.demo.prep, predictions_LP=="Low") %>% group_by(race) %>%
  summarize(count=n(), perce = (n()/nrow(filter(., predictions_LP=="Low")))*100)  
demo_racehigh = filter(df.demo.prep, predictions_LP=="High") %>% group_by(race) %>%
  summarize(count=n(), perce = (n()/nrow(filter(., predictions_LP=="High")))*100)  

## Assemble into readable matrix
demoRace = matrix(c(
  "White", demo_race$count[5], round(demo_race$perce[5],1) , demo_racelow$count[5], round(demo_racelow$perce[5],1), demo_racehigh$count[5], round(demo_racehigh$perce[5],1),   
  "Black", demo_race$count[2], round(demo_race$perce[2],1) , demo_racelow$count[2], round(demo_racelow$perce[2],1), demo_racehigh$count[2], round(demo_racehigh$perce[2],1),  
  "Hispanic", demo_race$count[3], round(demo_race$perce[3],1) , demo_racelow$count[3], round(demo_racelow$perce[3],1), demo_racehigh$count[3], round(demo_racehigh$perce[3],1),  
  "Asian", demo_race$count[1], round(demo_race$perce[1],1) , demo_racelow$count[1], round(demo_racelow$perce[1],1), demo_racehigh$count[1], round(demo_racehigh$perce[1],1),  
  "Other", demo_race$count[4], round(demo_race$perce[4],1) , demo_racelow$count[4], round(demo_racelow$perce[4],1), demo_racehigh$count[4], round(demo_racehigh$perce[4],1) ),
ncol=7,byrow=TRUE)
colnames(demoRace) = c("Levels", "Total", "Total-%", "LowHL", "LowHL-%", "HighHL", "HighHL-%")

pander(demoRace)

```

```{r}

## Get mean and SD for continuous variable of age; Assemble into readable matrix
demo_low = filter(df.demo.prep, predictions_LP=="Low") 
demo_high = filter(df.demo.prep, predictions_LP=="High") 

demoAge = matrix(c(
  "Age", round(mean(df.demo.prep$Age),1), round(sd(df.demo.prep$Age),1), round(mean(demo_low$Age),1), round(sd(demo_low$Age),1), round(mean(demo_high$Age),1), round(sd(demo_high$Age),1) ), 
  ncol=7,byrow=TRUE)
colnames(demoAge) = c("Levels", "Total", "Total-%", "LowHL", "LowHL-%", "HighHL", "HighHL-%")

pander(demoAge)

```

```{r}

## Get mean and SD for interval variable of comorbidity score; Assemble into readable matrix
democoMorbid = matrix(c(
  "Comorbidity Score", round(mean(df.demo.prep$CHARLSON_INDEX),1), round(sd(df.demo.prep$CHARLSON_INDEX),1), round(mean(demo_low$CHARLSON_INDEX),1), round(sd(demo_low$CHARLSON_INDEX),1), round(mean(demo_high$CHARLSON_INDEX),1), round(sd(demo_high$CHARLSON_INDEX),1) ), 
  ncol=7,byrow=TRUE)
colnames(democoMorbid) = c("Levels", "Total", "Total-%", "LowHL", "LowHL-%", "HighHL", "HighHL-%")

pander(democoMorbid)
```

#### Step 5: Simple tests to compare counts between high and low health literacy values

```{r}
library(compareGroups)

df.demo2 = df.demo

df.demo2$sex = as.factor(df.demo2$sex)
df.demo2$race = as.factor(df.demo2$race)
df.demo2$educ_p = as.factor(df.demo2$educ_p)

demo.test <- compareGroups(predictions_LP ~ Age + sex + race + educ_p + CHARLSON_INDEX,
    data = df.demo2)
print(demo.test)
```


