---
title: "Precision communication: Physicians’ linguistic adaptation to patients’ health literacy"
author: "Nick Duran"
date: 10/14/21
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
```

## What this Notebook Contains

> Notebook contains the complete statistical code for replicating section: 
> 
> * "Grouping by dyadic-level SM concordance"
> * "Association between dyadic-level SM concordance and patient-reported understanding"

#### Step 1: Import "Duran_Schillinger_KNPC_restricted_models.csv." 

See README in repository for instructions on how to access the .csv file. 

**Please note:** *Kaiser Permanante (owners of data) will not allow any individual-level data (even a small subset) to be shown in a public repository. A version of this HTML exists with the restricted .csv files that includes a subset of individual-level data to aid in tutorial interpretation.*

```{r, message=FALSE}
library(tidyverse)
library(pander)
```

```{r}
df.models = read.csv("Duran_Schillinger_KNPC_restricted_models.csv")

## Make sure to remove any patients who did not provide a CAHPS score for their physician
df.models = df.models[!(is.na(df.models$DV_CAREEXPL)),]
pander(names(df.models))
```

- **PCP_ID1**: Randomly generated numeric code associated with each unique doctor 
- **predictions_LP**: Patient's assigned score of "high" or "low" health literacy (otherwise known as "LP Score") based on previously developed linguistic model; see: S. A. Crossley, R. Balyan, J. Liu, A. J. Karter, D. McNamara, D. Schillinger, Developing and testing automatic models of patient communicative health literacy using linguistic features: Findings from the ECLIPPSE study, Health Commun. (2020)
- **predictions_new_CP**: The doctor's score of "high" or "low" writing complexity (for a particular patient; otherwise known as "CP Score") based on previously developed linguistic model developed; see S. A. Crossley, R. Balyan, J. Liu, A. J. Karter, D. McNamara, D. Schillinger, Predicting the readability of physicians’ secure messages to improve health communication using novel linguistic features: Findings from the ECLIPPSE study, J. Commun. Healthc. 13, 344–356 (2020). 
- **DV_CAREEXPL**: Doctor's CAHPS score received by a patient
- **cat_match**: Doctor's CP (Complexty Profile) score associated with a patient's LP (Health Literacy) score, basis for determining language concordance/matching (e.g., four possibilities: lowCP-lowLP, lowCP-highCP, highCP-lowLP, highCP-highCP)
- **Age_z**: Age of patient (z-scored)
- **sex**: Sex of patient; 0 = female, 1 = male
- **white**: Race of patient; 1 = White, 0 = Non-White
- **ed_HS**: Education level of patient; 1 = Some College, 2 = No College
- **CHARLSON_INDEX_z**: Co-morbidity index, interval (z-scored)

## Subhead: "Grouping by dyadic-level SM concordance"

#### Step 1: Create new variable "cat_match" that captures the concordance relationship between patient and physician based on respective linguistic score (low or high)

```{r}
df.models = df.models %>% mutate(
  cat_match = case_when(
    predicted_new_CP=="low" & predictions_LP=="Low" ~ "lowCP-lowLP",
    predicted_new_CP=="low" & predictions_LP=="High" ~ "lowCP-highLP",
    predicted_new_CP=="high" & predictions_LP=="Low" ~ "highCP-lowLP",
    predicted_new_CP=="high" & predictions_LP=="High" ~ "highCP-highLP",
    TRUE ~ NA_character_))
    
df.models$cat_match = as.factor(df.models$cat_match)
# pander(names(df.models[c(1,3,4,10)]))
```

#### Step 2: Assign whether the CAHPS survey score provided by patient about physician was "good" or "poor" and show breakdown of score based on concordance relationship

```{r}
df.models.tab = df.models %>%
  mutate(DV_CAREEXPL = case_when(
    DV_CAREEXPL=="0" ~ "Good", 
    DV_CAREEXPL=="1" ~ "Poor"))
```

```{r}
x1 = xtabs(~ DV_CAREEXPL + cat_match, df.models.tab)
pander(x1)
```
#### Step 3: Generate summary descriptives as described in manuscript

"Across all 4,331 physician/patient dyads..."

```{r}
pander(sum(x1))
```

"the prevalence of discordance (47.19%; 2,044 pairs) was lower than concordance (52.81%, 2,287 pairs)."

```{r}
pander(sum(rowSums(t(x1))[c(2,3)]))
pander(sum(rowSums(t(x1))[c(1,4)]))
```

"Stratifying by patients’ HL, the proportional frequency of discordance (relative to concordance) was higher amongst patients with low HL compared to high HL (p<.001). Of the 1,560 patients classified as low HL, 821 (52.62%) had a discordant physician (i.e., high complexity), whereas among the 2,771 patients classified as high HL, 1,223 (44.14%) had a discordant physician (i.e., low complexity)."

```{r}
pander(prop.test(x = c(1223, 821), n = c(2771, 1560)))
```

## Subhead: "Association between dyadic-level SM concordance and patient-reported understanding" 

#### Step 1: Generate summary descriptives as described in manuscript.

"Overall, 475 (10.97%) of all patients reported “poor” understanding of their healthcare provider..."

```{r}
pander(colSums(t(x1)))
```

"...this lack of understanding was more prevalent among low vs. high HL patients (13.65% vs 9.46%, p<.001)."

```{r}
pander(prop.test(x = c(sum(x1[2,c(1,3)]), sum(x1[2,c(2,4)])), n = c(sum(x1[,c(1,3)]), sum(x1[,c(2,4)]))))
```

### Mixed effects models

**Corresponds to results reported in "Table 2. Results of dyadic-level concordance and patient-reported understanding**

```{r, warning=FALSE, message=FALSE, error=FALSE}
library(lme4)
library(strengejacke)
```

```{r}
## Standardize appropriate variables
df.models.var = df.models %>% mutate(
                                 CHARLSON_INDEX_z = scale(CHARLSON_INDEX, center=FALSE),
                                 Age_z = scale(Age, center=FALSE)
                                 )
```

#### Focus on Low HL patients

```{r}
df_low = df.models.var %>% filter(cat_match == "lowCP-lowLP" | cat_match == "highCP-lowLP")
df_low$cat_match <- relevel(df_low$cat_match, "lowCP-lowLP")

glmer.fit1 <- glmer(DV_CAREEXPL ~ cat_match + Age_z + sex + white + CHARLSON_INDEX_z + ed_HS + (1|PCP_ID1), df_low, family = binomial)
tab_model(glmer.fit1,digits = 3,digits.re = 3)
```

#### Focus on High HL patients

```{r}
df_high = df.models.var %>% filter(cat_match == "lowCP-highLP" | cat_match == "highCP-highLP")
df_high$cat_match <- relevel(df_high$cat_match, "highCP-highLP")

glmer.fit2 <- glmer(DV_CAREEXPL ~ cat_match + Age_z + sex + white + CHARLSON_INDEX_z + ed_HS + (1|PCP_ID1), df_high, family = binomial)
tab_model(glmer.fit2,digits = 3,digits.re = 3)
```


















